Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-22 ms.tex
CP's final changes before the internal referee process

\begin{abstract}
  Observations of radio halos and relics in galaxy clusters are indicative of efficient acceleration
  processes of electrons while this process should be even more efficient for protons. In
  combination with their long cooling time which exceeds the Hubble time, this suggests they may also
  be sources of very high-energy (VHE) gamma-ray emission as a result of highly relativistic
  cosmic-ray (CR) protons and ions interacting with intracluster gas. Reported here are VHE
  gamma-ray observations of the Coma galaxy cluster with the VERITAS array of imaging Cherenkov
  telescopes, with complementing Fermi Gamma-ray Space Telescope observations at GeV energies. No
  significant gamma-ray emission from the Coma cluster was detected. Flux upper limits at the 99\%
  confidence level were measured to be on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$
  (VERITAS, $>$220 GeV) and $\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$ ({\em Fermi}, 1-3 GeV),
  respectively. For {\em Fermi} (VERITAS), the limits are a factor of 20 (60) higher than
  model-predicted minimum gamma-ray fluxes, which assume the radio-emitting electrons to be all
  secondaries from cosmic-ray interactions. We use the gamma-ray flux upper limits to constrain the
  populations of CRs and magnetic fields and in Coma. Using an analytical approach informed by
  cosmological cluster simulations, the CR-to-thermal pressure is constrained to be smaller than
  15\% from VERITAS data and smaller than 2.5\% from {\em Fermi} data (averaged within the virial
  radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic CR
  interactions, the observations imply a lower limit on the intra-cluster magnetic field strengths
  of about $1$--$2\mu$G, depending on the magnetic decline and gamma-ray spectral index.  Since
  these values are well below those inferred by Faraday rotation measurements in Coma, this renders
  the hadronic model a very plausible explanation of the Coma radio halo. Finally, a constraint on
  the thermally averaged product of the total self-annihilation cross section and the relative
  velocity of the dark-matter particles, $\expval{\sigma v}$, is derived and found to be $\lesssim
  10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}

@@

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung in the soft
X-rays regime \citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated
primarily through collisionless structure-formation shocks that form as a result of the
hierarchical merging and accretion processes. Such shocks and turbulence in the ICM gas and
intra-cluster magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two theories competing to explain the radio halo
are the ``hadronic model'', in which the radio-emitting electrons are produced in inelastic
collisions of cosmic-ray (CR)  ions with thermal gas of the ICM \citep{article:Dennison:1980},
%\citep[][and references therein]{article:EnsslinPfrommerMiniatiSubramanian:2011}
and the ``re-acceleration model'', in which during states of strong ICM turbulence, e.g. after a
cluster merger, interactions of a long-lived pool of 100-MeV electrons with plasma waves might
be efficient enough to accelerate them to energies ($\sim 10$ GeV) sufficient to produce observable
radio emission \citep[][and references therein]{article:BrunettiLazarian:2010}. Recent observations of
possibly non-thermal emission from clusters in the the extreme ultraviolet
\citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays \citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may provide further indications of
relativistic particle populations in clusters, although the interpretation of these observations as non-thermal diffuse emission have been disputed on the basis of more sensitive observations
\citep[see, e.g.,][]{article:Ajello_etal:2009, article:Ajello_etal:2010, article:Wik_etal:2009}.

Galaxy clusters have for many years been proposed as sources of gamma rays. If shock acceleration
in the ICM is an efficient process, a population of highly relativistic CR protons and heavy ions is to
be expected in the ICM. The main energy-loss mechanism for CR hadrons at high energies is pion
production through the interaction of CRs with nuclei in the ICM. Pions are short lived and decay; the
charged pions to electrons and positrons,\footnote{A charged pion decays to a muon or antimuon,
which subsequently decay to an electron or positron as well as neutrinos.} and the neutral pion to
gamma rays. Due to the low density of the ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large
scales of a cluster and the volume filling magnetic fields in the ICM, CR hadrons will be confined in
the cluster on timescales comparable to or longer than the Hubble time \citep[][]{article:Volk_etal:1996,
article:Berezinsky_etal:1997} allowing them to accumulate over time. Depending on the CR pressure,
quantified by the CR-to-thermal pressure fraction
$X_\CR=\expval{P_{\CR}}/\expval{P_{\mathrm{th}}}$, an observable flux of high-energy gamma rays
can be expected.

A substantial value of $X_\CR$ can bias hydrostatic estimates of cluster masses as well as the
temperature decrement of the cosmic microwave background (CMB) in the direction of a galaxy cluster
due to the Sunyaev-Zel'dovich effect. This could then severely jeopardize the use of clusters to
determine cosmological parameters. Comparing X-ray and optical potential profiles in the centers of
galaxy clusters yields an upper limit of 20-30 \% of non-thermal pressure (that can be composed of
CRs, magnetic fields or turbulence) relative to the thermal gas pressure
\citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially
resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters
detects a deficit of the hydrostatic mass estimate compared to the lensing mass of $20 \%$ at
$R_{500}$ -- the radius within which the mean density is 500 times the critical density --
suggesting again a substantial non-thermal pressure contribution on large scales
\citep{article:Mahdavi_etal:2008}. Observing high-energy gamma-ray emission is a complementary
method of constraining the pressure contribution of CRs that is most sensitive to the cluster core
region. However, it assumes that the CR component is fully mixed with the ICM and may not a
two-phase structure of CRs and the thermal ICM. A fraction $X_\CR$ of only a few percent is
required in order to produce a gamma-ray flux observable with current generation gamma-ray
telescopes rendering this technique at least as sensitive as the dynamical and hydrostatic method
(which are more general in that they are sensitive to any non-thermal pressure component).

@@

While several observations of clusters of galaxies have been made with satellite-borne and
ground-based gamma-ray telescopes, a detection of gamma-ray emission from a cluster has yet to be
made. Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the
Fermi Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on the
gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for Fermi observations) for several
galaxy clusters in the MeV to GeV band. Upper limits on the very-high-energy (VHE) gamma-ray flux
from a few select clusters, including the Coma cluster, have been provided by observations with
ground-based imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Perkins_etal:2006,
inproc:Perkins_etal:2008, article:Aharonian_etal:2009a, article:Aleksic_etal:2010}.

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied clusters over all
wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc
\citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest massive clusters
\citep[$M \sim10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a
giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral
radio relic, which appears connected to the radio halo with a ``diffuse'' bridge 
\citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested
\citep{article:Ensslin_etal:1998} and successively demonstrated by cosmological simulations, which
model the non-thermal emission processes, \citep{article:PfrommerEnsslinSpringel:2008,
article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}, that the relic is
an infall shock. Extended soft thermal X-ray (SXR) emission is evident from the ROSAT all-sky
survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992}. Observations with
XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo supported by
substantial turbulent pressure of at least $\sim 10 \%$ of the total pressure
\citep{article:Schuecker_etal:2004}. The Coma cluster is a natural candidate for gamma-ray
observations.

@@

The data analysis was performed following the standard VERITAS procedures described in
\citet{inproc:Cogan_etal:2007} and \citet{inproc:Daniel_etal:2007}. Prior to event reconstruction
and selection, all shower images are calibrated and cleaned. Showers are then reconstructed for
events with at least two contributing telescope images that pass the following quality selection
criteria: more than four participating pixels in the camera, number of photoelectrons in the image
is larger than 75, and the distance from the image centroid to the center of the camera is less
than $1.43^{\circ}$. These quality selection criteria impose an energy threshold\footnote{Defined
as the energy corresponding to the maximum of the product function of the observed spectrum with
the collection area} of about 220 GeV. In addition, events for which only images from the two
closest-spaced telescopes\footnote{In the array configuration prior to summer 2009, two telescopes 
had a separation of only 35 m.} survive quality selection are rejected, as they introduce an
irreducible high background rate, due to local muons, degrading the instrument sensitivity
\citep{article:MaierKnapp:2007}.

@@

Table \ref{table:results} lists the upper limits for the selected regions of interest tabulated in
Table \ref{table:roi}. These upper-limit calculations depend on the gamma-ray spectrum, which in
this work is assumed to be a power law with a spectral index $\alpha=2.1, 2.3$, and $2.5$.

@@

The analysis of this work follows the same procedure as described in detail in
\citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle
cut of 105$^{\circ}$ is applied. The \emph{Fermi} Science Tools version v9r23p1 and Instrumental
Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events
calibrated  in flight) were used throughout this work. Only photons above 1 GeV are considered,
where the constraints on the flux are most meaningful. The \emph{Fermi}-LAT collaboration estimates
the systematic uncertainties at 10 GeV to be around 20\%.
\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/LAT\_caveats.html}} Flux upper limits
at the 99\% confidence level in the energy range 1--30 GeV, assuming both a unresolved, point-like
or spatially extended emission, are tabulated in Table \ref{table:fermi}. The gamma-ray emission is
assumed to follow a power-law distribution with a photon index $\alpha=2.1$, 2.3, and $2.5$.

@@

We start by adopting a simplified analytical model that assumes a power-law CR spectrum and a
constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs} following the approach
of \citet{article:PfrommerEnsslin:2004b}. To be independent of additional assumptions and in line
with earlier work in the literature, we do not impose a low-momentum cutoff on the CR distribution
function, i.e., we adopt $q=0$. Since a priori, the CR spectral index is unconstrained, we vary it in
the range $2.1<\alpha_{\rmn{p}}<2.5$, which is compatible with the
radio spectral index of the giant radio halo of the Coma cluster after accounting for the spectral
steepening at frequencies $\nu\sim5~\rmn{GHz}$ due to the Sunyaev-Zel'dovich effect
\citep{article:PfrommerEnsslin:2004b}.\footnote{Assuming a magnetic field of 1 $\mu$G, the CR protons
responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are $\sim$ 20
times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.} To model
the thermal pressure, we adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and use a constant
temperature of $kT= 8.25$~keV throughout the virial region.

Table \ref{table:constraints_simple} shows the resulting constraints on the relative CR pressure, $X_{\CR}
= \expval{P_{\CR}}/\expval{P_\rmn{th}}$, averaged within the virial radius, $R_\rmn{vir}=2.2$~Mpc,
that we define as the radius of a sphere enclosing a mean density that is 200 times the critical
density of the Universe. The VERITAS flux upper limits strongly depend on $\alpha$ due to the
comparably large lever arm from GeV energies (that dominate the CR pressure) to TeV-energies. The
flux measurements within 0.2$^{\circ}$ are the most constraining and yield limits on $X_\CR$
between 0.048 and 0.43
(for $\alpha$ varying between 2.1 and 2.5), with a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close
to the spectral index predicted by the simulations of \citet{article:PinzkePfrommer:2010} around 220
GeV).  The {\em Fermi} limits depend only weakly on $\alpha$ and are most constraining for an
aperture of 0.4$^{\circ}$.  The best limit of $X_\CR< 0.017$ is achieved for $\alpha=2.3$ and only
slightly worse for $\alpha\simeq 2.1$ yielding $X_\CR<0.025$.

@@

The overall normalization of the CR and gamma-ray distribution scales non-linearly with the
acceleration efficiency at structure formation shock. Following recent observations at
supernova remnants \citep{article:Helder_etal:2009} as well as theoretical studies
\citep{article:KangJones:2005}, we adopt an optimistic but nevertheless realistic value of this
parameter and assume that 50\% of the dissipated energy at strong shocks is injected into CRs while
this efficiency rapidly decreases for weaker shocks. Since the vast majority of internal formation
shocks (merger and flow shocks) are weak shocks with Mach numbers $M\lesssim3$
\citep[e.g.,][]{article:Ryu_etal:2003}, they do not significantly contribute to the CR population in
clusters. Instead, strong shocks during the formation epoch of clusters and strong accretion shocks
at the present time (at the boundary of voids and filaments/super-cluster regions) dominate the
injection of CRs which subsequently are adiabatically transported through the cluster. Hence, the model
provides a plausible upper limit for the CR contribution from structure formation shocks
in galaxy clusters which can be scaled with the effective acceleration efficiency. Other possible CR
sources, such as AGN and starburst-driven galactic winds have been neglected for simplicity which
could in principle increase the expected gamma-ray yield.

@@

In the case of VERITAS, for the most constraining regions tested (within an aperture of radius
0.2$^{\circ}$), the predicted CR pressure is a factor of 7.2 below the inferred upper limits of VERITAS
(see Table
\ref{table:results} and assuming a spectral index of $\alpha=2.3$ which matches the simulated one at
energies $E_\gamma=200$ GeV). To first order, we can scale the averaged relative CR pressure of
our model by that
factor, keep the spatial behavior and obtain an integrated limit of the relative CR pressure of
$X_{\CR}<0.105$ within 0.2$^{\circ}$ that translates to a limit within the cluster virial radius of
$X_\CR<0.152$ (solid lines of Figure \ref{fig:xcr}).  This limit is less constraining by 50\% in
comparison to the simplified analytical model, which gives $X_\CR<0.1$. This difference is
explained by the concave curvature of the simulated spectrum which accumulates additional pressure
toward GeV energies in comparison to a pure power-law spectrum of $\alpha=2.3$.

@@

\subsection{Minimum gamma-ray flux}
\label{sec:Fmin}
For clusters that host radio halos we can derive a minimum gamma-ray flux in the hadronic
model of radio halos -- where the radio-emitting electrons are secondaries from CR interactions. 
Hadronic interactions channel about the same power into secondary electrons and $\pi^{0}$-decay
gamma rays. A stationary distribution of CR electrons loses all its energy to
synchrotron radiation for strong magnetic fields ($B \gg B_{\rmn{CMB}} \simeq 3.2 \mu$G), and thus the
ratio of gamma-ray to synchrotron flux becomes independent of the spatial distribution of CRs and
thermal gas \citep{article:Voelk:1989, article:Pohl:1994, article:Pfrommer:2008}, in particular with
$\alpha_{\nu}\simeq 1$ as the observed synchrotron spectral index.  Hence we can derive a minimum
gamma-ray flux in the hadronic model
\begin{equation}
\label{eq:Fmin}
\mathcal{F}_{\gamma,\rmn{min}} = \frac{\dps A_{\gamma}}{\dps A_{\nu}}\frac{\dps L_{\nu}}{\dps 4\pi D_{\rmn{lum}}^{2}},
\end{equation}
where $L_{\nu}$ is the observed luminosity of the radio mini-halo, $D_{\rmn{lum}}$ denotes the
luminosity distance to the respective cluster, $A_\gamma$ and $A_\nu$ are dimensional constants
that depend on the hadronic physics of the interaction \citep{article:Pfrommer:2008,
Pfrommer_etal:2008}. Lowering the magnetic field would require an increase in the energy density of
CR electrons to reproduce the observed synchrotron luminosity and thus increase the associated
gamma-ray flux.

@@

\item
We then solve for $B_{0}$ while matching the observed synchrotron profile while fixing the profile
of $X_{\CR}(r)$ as determined through the previous two steps. Note
that for $B_{0} \gg B_{\rmn{CMB}}$ and a radio spectral index of $\alpha_{\nu}=1$, the solution
would be degenerate.

>>>>

\begin{abstract}
  Observations of radio halos and relics in galaxy clusters are indicative of efficient acceleration
  processes of electrons while this process should be even more efficient for protons. In
  combination with their long cooling time which exceeds the Hubble time, this suggests they may
  also be sources of very high-energy (VHE) gamma-ray emission as a result of highly relativistic
  cosmic-ray (CR) protons and ions interacting with intracluster gas. Reported here are VHE
  gamma-ray observations of the Coma galaxy cluster with the VERITAS array of imaging Cherenkov
  telescopes, with complementing Fermi Gamma-ray Space Telescope observations at GeV energies. No
  significant gamma-ray emission from the Coma cluster was detected. Flux upper limits at the 99\%
  confidence level were measured to be on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$ s$^{-1}$
  (VERITAS, $>$220 GeV) and $\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$ ({\em Fermi}, 1-3 GeV),
  respectively. For VERITAS ({\em Fermi}), the limits are a factor of 60 (20) higher than
  model-predicted minimum gamma-ray fluxes, which assume the radio-emitting electrons of the Coma
  radio halo to be all secondaries from cosmic-ray interactions. We use the gamma-ray flux upper
  limits to constrain the populations of CRs and magnetic fields and in Coma. Using an analytical
  approach informed by cosmological cluster simulations, the CR-to-thermal pressure is constrained
  to be $< 15\%$ from VERITAS data and $< 2.5\%$ from {\em Fermi} data (averaged within the virial
  radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic CR
  interactions, the observations imply a lower limit on the central magnetic field in Coma of
  $0.5$--$3\mu$G, depending on the magnetic decline and gamma-ray spectral index.  Since these
  values are well below those inferred by Faraday rotation measurements in Coma, this makes the
  hadronic model a very plausible explanation of the Coma radio halo. Finally, a constraint on the
  thermally averaged product of the total self-annihilation cross section and the relative velocity
  of the dark-matter particles, $\expval{\sigma v}$, is derived and found to be $\lesssim 10^{-21}$
  cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}

@@

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung in the soft X-ray regime
\citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated primarily through
collisionless structure-formation shocks that form as a result of the hierarchical merging and
accretion processes. Such shocks and turbulence in the ICM gas in combination with intra-cluster
magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two theories competing to explain the radio halo are the
``hadronic model'', in which the radio-emitting electrons are produced in inelastic collisions of
cosmic-ray (CR) ions with thermal gas of the ICM \citep{article:Dennison:1980,
  article:EnsslinPfrommerMiniatiSubramanian:2011}, and the ``re-acceleration model'', in which
during states of strong ICM turbulence, e.g. after a cluster merger, interactions of a long-lived
pool of 100-MeV electrons with plasma waves might be efficient enough to accelerate them to energies
($\sim 10$ GeV) sufficient to produce observable radio emission \citep[][and references
therein]{article:BrunettiLazarian:2010}. Recent observations of possibly non-thermal emission from
clusters in the the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays
\citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may
provide further indications of relativistic particle populations in clusters, although the
interpretation of these observations as non-thermal diffuse emission have been disputed on the basis
of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
  article:Ajello_etal:2010, article:Wik_etal:2009}.

Galaxy clusters have for many years been proposed as sources of gamma rays. If shock acceleration
in the ICM is an efficient process, a population of highly relativistic CR protons and heavy ions is to
be expected in the ICM. The main energy-loss mechanism for CR hadrons at high energies is pion
production through the interaction of CRs with nuclei in the ICM. Pions are short lived and decay; the
charged pions to electrons and positrons,\footnote{A charged pion decays to a muon or antimuon,
which subsequently decay to an electron or positron as well as neutrinos.} and the neutral pion to
gamma rays. Due to the low density of the ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large
scales of a cluster and the volume-filling magnetic fields in the ICM, CR hadrons will be confined in
the cluster on timescales comparable to or longer than the Hubble time \citep[][]{article:Volk_etal:1996,
article:Berezinsky_etal:1997} allowing them to accumulate over time. Depending on the CR pressure,
quantified by the CR-to-thermal pressure fraction
$X_\CR=\expval{P_{\CR}}/\expval{P_{\mathrm{th}}}$, an observable flux of high-energy gamma rays
can be expected.

A substantial value of $X_\CR$ can bias hydrostatic estimates of cluster masses as well as the
temperature decrement of the cosmic microwave background (CMB) in the direction of a galaxy cluster
due to the Sunyaev-Zel'dovich effect. This could then severely jeopardize the use of clusters to
determine cosmological parameters. Comparing X-ray and optical potential profiles in the centers of
galaxy clusters yields an upper limit of 20-30 \% of non-thermal pressure (that can be composed of
CRs, magnetic fields or turbulence) relative to the thermal gas pressure
\citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially
resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters
detects a deficit of the hydrostatic mass estimate compared to the lensing mass of $20 \%$ at
$R_{500}$ -- the radius within which the mean density is 500 times the critical density --
suggesting again a substantial non-thermal pressure contribution on large scales
\citep{article:Mahdavi_etal:2008}. Observing high-energy gamma-ray emission is a complementary
method of constraining the pressure contribution of CRs that is most sensitive to the cluster core
region. However, it assumes that the CR component is fully mixed with the ICM and may not allow to
detect a two-phase structure of CRs and the thermal ICM. A fraction $X_\CR$ of only a few percent is
required in order to produce a gamma-ray flux observable with current generation gamma-ray
telescopes rendering this technique at least as sensitive as the dynamical and hydrostatic method
(which are more general in that they are sensitive to any non-thermal pressure component).

@@

While several observations of clusters of galaxies have been made with satellite-borne and
ground-based gamma-ray telescopes, a detection of gamma-ray emission from a cluster has yet to be
made. Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the
Fermi Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on the
gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for Fermi observations) for several
galaxy clusters in the MeV to GeV band. Upper limits on the very-high-energy (VHE) gamma-ray flux
from a few selected clusters, including the Coma cluster, have been provided by observations with
ground-based imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Perkins_etal:2006,
inproc:Perkins_etal:2008, article:Aharonian_etal:2009a, article:Aleksic_etal:2010}.

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied clusters over all
wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc
\citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest massive clusters \citep[$M
\sim10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio
halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic,
which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion
in][]{article:BrownRudnick:2010}. It has been suggested \citep{article:Ensslin_etal:1998} and
successively demonstrated by cosmological simulations, which model the non-thermal emission
processes, \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008,
  article:Battaglia_etal:2009, article:Skillman_etal:2011}, that the relic could well be an infall
shock. Extended soft thermal X-ray (SXR) emission is evident from the ROSAT all-sky survey in the
0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992}. Observations with XMM-Newton
\citep{article:Briel_etal:2001} revealed substructure in the X-ray halo supported by substantial
turbulent pressure of at least $\sim 10 \%$ of the total pressure
\citep{article:Schuecker_etal:2004}. The Coma cluster is a natural candidate for gamma-ray
observations.

@@

The data analysis was performed following the standard VERITAS procedures described in
\citet{inproc:Cogan_etal:2007} and \citet{inproc:Daniel_etal:2007}. Prior to event reconstruction
and selection, all shower images are calibrated and cleaned. Showers are then reconstructed for
events with at least two contributing telescope images that pass the following quality selection
criteria: more than four participating pixels in the camera, number of photoelectrons in the image
is larger than 75, and the distance from the image centroid to the center of the camera is less
than $1.43^{\circ}$. These quality selection criteria impose an energy threshold\footnote{Defined
as the energy corresponding to the maximum of the product function of the observed spectrum with
the collection area.} of about 220 GeV. In addition, events for which only images from the two
closest-spaced telescopes\footnote{In the array configuration prior to summer 2009, two telescopes 
had a separation of only 35 m.} survive quality selection are rejected, as they introduce an
irreducible high background rate, due to local muons, degrading the instrument sensitivity
\citep{article:MaierKnapp:2007}.

@@

Table \ref{table:results} lists the upper limits for the selected regions of interest tabulated in
Table \ref{table:roi}. These upper-limit calculations depend on the gamma-ray spectrum, which in
this work is assumed to be a power law in energy, $F\propto E^{-\alpha}$, with a spectral index
$\alpha=2.1, 2.3$, and $2.5$.

@@

The analysis of this work follows the same procedure as described in detail in
\citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle
cut of 105$^{\circ}$ is applied. The \emph{Fermi} Science Tools version v9r23p1 and Instrumental
Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events
calibrated in flight) were used throughout this work. Only photons above 1 GeV are considered, where
the constraints on the flux are most meaningful. The \emph{Fermi}-LAT collaboration estimates the
systematic uncertainties at 10 GeV to be around
20\%.\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/LAT\_caveats.html}} Flux upper
limits at the 99\% confidence level in the energy range 1--30 GeV, assuming both a unresolved,
point-like or spatially extended emission, are tabulated in Table \ref{table:fermi}. The gamma-ray
emission is assumed to follow a power-law distribution with a photon index $\alpha=2.1$, 2.3, and
$2.5$.

@@

We start by adopting a simplified analytical model that assumes a power-law CR spectrum and a
constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs} following the approach
of \citet{article:PfrommerEnsslin:2004b}. To be independent of additional assumptions and in line
with earlier work in the literature, we do not impose a low-momentum cutoff on the CR distribution
function, i.e., we adopt $q=0$. Since a priori, the CR spectral index is unconstrained,\footnote{The
  hadronic interaction physics guarantees that the CR spectral index coincides with that of the
  resulting pion-decay gamma-ray emission.} we vary it in the range $2.1<\alpha<2.5$, which is
compatible with the radio spectral index of the giant radio halo of the Coma cluster after
accounting for the spectral steepening at frequencies $\nu\sim5~\rmn{GHz}$ due to the
Sunyaev-Zel'dovich effect \citep{article:Ensslin:2002,
  article:PfrommerEnsslin:2004b}.\footnote{Assuming a magnetic field of 1 $\mu$G, the CR protons
  responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are $\sim$ 20
  times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.} To model
the thermal pressure, we adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and use a constant
temperature of $kT= 8.25$~keV throughout the virial region.

Table \ref{table:constraints_simple} shows the resulting constraints on the relative CR pressure,
$X_{\CR} = \expval{P_{\CR}}/\expval{P_\rmn{th}}$, averaged within the virial radius,
$R_\rmn{vir}=2.2$~Mpc, that we define as the radius of a sphere enclosing a mean density that is 200
times the critical density of the Universe. Constraints on $X_\CR$ with VERITAS flux upper limits
strongly depend on $\alpha$ due to the comparably large lever arm from GeV energies (that dominate
the CR pressure) to TeV energies. The flux measurements within 0.2$^{\circ}$ are the most
constraining and yield limits on $X_\CR$ between 0.048 and 0.43 (for $\alpha$ varying between 2.1
and 2.5), with a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close to the spectral index predicted
by the simulations of \citet{article:PinzkePfrommer:2010} around 220 GeV).  Constraints on $X_\CR$
with {\em Fermi} limits depend only weakly on $\alpha$ and are most constraining for an aperture of
0.4$^{\circ}$.  The best limit of $X_\CR< 0.017$ is achieved for $\alpha=2.3$ and only slightly
worse for $\alpha\simeq 2.1$ yielding $X_\CR<0.025$.

@@

The overall normalization of the CR and gamma-ray distribution scales non-linearly with the
acceleration efficiency at structure formation shocks. Following recent observations at supernova
remnants \citep{article:Helder_etal:2009} as well as theoretical studies
\citep{article:KangJones:2005}, we adopt an optimistic but nevertheless realistic value of this
parameter and assume that 50\% of the dissipated energy at strong shocks is injected into CRs while
this efficiency rapidly decreases for weaker shocks. Since the vast majority of internal formation
shocks (merger and flow shocks) are weak shocks with Mach numbers $M\lesssim3$
\citep[e.g.,][]{article:Ryu_etal:2003}, they do not significantly contribute to the CR population in
clusters. Instead, strong shocks during the formation epoch of clusters and strong accretion shocks
at the present time (at the boundary of voids and filaments/super-cluster regions) dominate the
acceleration of CRs which subsequently are adiabatically transported through the cluster. Hence, the
model provides a plausible upper limit for the CR contribution from structure formation shocks in
galaxy clusters which can be scaled with the effective acceleration efficiency. Other possible CR
sources, such as AGN and starburst-driven galactic winds have been neglected for simplicity which
could in principle increase the expected gamma-ray yield.

@@

In the case of VERITAS, for the most constraining regions tested (within an aperture of radius
0.2$^{\circ}$), the predicted CR pressure is a factor of 7.2 below the inferred upper limits of
VERITAS (see Table \ref{table:results} and assuming a spectral index of $\alpha=2.3$ which matches
the simulated one at energies $E_\gamma=200$ GeV). To first order, we can scale the averaged
relative CR pressure of our model by that factor, keep the spatial behavior and obtain an integrated
limit of the relative CR pressure of $X_{\CR}<0.105$ within 0.2$^{\circ}$ that translates to a limit
within the cluster virial radius of $X_\CR<0.152$ (solid lines of Figure \ref{fig:xcr}).  This limit
is less constraining by 50\% in comparison to the simplified analytical model, which gives
$X_\CR<0.1$. This difference is explained by the concavity of the simulated spectrum which therefore
carries more pressure at GeV energies than a pure power-law spectrum of $\alpha=2.3$.

@@

\subsection{Minimum gamma-ray flux}
\label{sec:Fmin}
For clusters that host radio halos we can derive a minimum gamma-ray flux in the hadronic
model of radio halos -- where the radio-emitting electrons are secondaries from CR interactions. 
Hadronic interactions channel about the same power into secondary electrons and $\pi^{0}$-decay
gamma rays. A stationary distribution of CR electrons loses all its energy to
synchrotron radiation for strong magnetic fields ($B \gg B_{\rmn{CMB}} \simeq 3.2 \mu$G), and thus the
ratio of gamma-ray to synchrotron flux becomes independent of the spatial distribution of CRs and
thermal gas \citep{article:Voelk:1989, article:Pohl:1994, article:Pfrommer:2008}, in particular with
$\alpha_{\nu}\simeq 1$ as the observed synchrotron spectral index.  Hence we can derive a minimum
gamma-ray flux in the hadronic model
\begin{equation}
\label{eq:Fmin}
F_{\gamma,\rmn{min}} = \frac{\dps A_{\gamma}}{\dps A_{\nu}}\frac{\dps L_{\nu}}{\dps 4\pi D_{\rmn{lum}}^{2}},
\end{equation}
where $L_{\nu}$ is the observed luminosity of the radio mini-halo, $D_{\rmn{lum}}$ denotes the
luminosity distance to the respective cluster, $A_\gamma$ and $A_\nu$ are dimensional constants
that depend on the hadronic physics of the interaction \citep{article:Pfrommer:2008,
Pfrommer_etal:2008}. Lowering the magnetic field would require an increase in the energy density of
CR electrons to reproduce the observed synchrotron luminosity and thus increase the associated
gamma-ray flux.

@@

\item
We then solve for $B_{0}$ while matching the observed synchrotron profile and fixing the profile
of $X_{\CR}(r)$ as determined through the previous two steps. Note
that for $B_{0} \gg B_{\rmn{CMB}}$ and a radio spectral index of $\alpha_{\nu}=1$, the solution
would be degenerate.
